---
title: "Synthetic Data Generation and Evaluation for WP13 AT"
author: "WP13 AT"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
editor: visual
---

# Overview

This document demonstrates a compact workflow for generating synthetic versions of the `eusilcP` population data and comparing the resulting data sets with basic disclosure-risk and utility-oriented measures.

The workflow has four main parts:

1.  Install and load the required R and Python dependencies.
2.  Prepare the `eusilcP` population data.
3.  Generate synthetic data with `simPop`, `synthpop`, and a tabular ARGAN workflow.
4.  Evaluate the synthetic data with replicated uniques and RAPID.

The code chunks are shown but not evaluated by default. This avoids accidental package installation, Python environment changes, and long model-fitting steps when rendering the document. To run the full workflow, change `eval: false` to `eval: true` in the YAML header or run the chunks interactively.

# Environment Setup

The workflow depends on several R packages, one GitHub package, and a Python package used through `reticulate`. The installation steps are separated from the analysis so they can be run once per machine.

```{r}
#| label: install-r-packages

install.packages(c(
  "data.table",
  "simPop",
  "synthpop",
  "devtools",
  "reticulate",
  "keras"
))

devtools::install_github("qwertzlbry/RAPID")
keras::install_keras()
```

The tabular ARGAN part of the workflow uses the Python package `mostlyai`. The command below installs the package into the Keras virtual environment.

```{r}
#| label: install-python-packages

system("~/.virtualenvs/r-keras/bin/pip install -U 'mostlyai[local]'")
```

After installation, point `reticulate` to the same Python executable. In a fresh R session, this should be set before loading packages that initialize Python.

```{r}
#| label: configure-reticulate

Sys.setenv(RETICULATE_PYTHON = "~/.virtualenvs/r-keras/bin/python")
```

Load the R packages used throughout the example.

```{r}
#| label: load-packages

library(data.table)
library(simPop)
library(synthpop)
library(devtools)
library(RAPID)
```

# Data Preparation

The example uses the `eusilcP` population data from `simPop`. We convert it to a `data.table` and derive two total income variables:

- `IncomeP`: total personal income, calculated from variables starting with `py`
- `IncomeH`: total household income, calculated from variables starting with `hy`

```{r}
#| label: load-and-prepare-data

data(eusilcP)
setDT(eusilcP)

cnames <- colnames(eusilcP)
income_h <- cnames[cnames %like% "^hy"]

eusilcP[, IncomeP := rowSums(.SD), .SDcols = patterns("^py")]
eusilcP[, IncomeH := rowSums(.SD[1,]), by = .(hid), .SDcols = patterns("^hy")]
```

# Synthetic Data Generation

This section applies three different synthetic data generation approaches:

- `simPop`, using household structure and sequential variable simulation
- `synthpop`, using sequential synthesis over selected variables
- tabular ARGAN, using a Python-based generation workflow through `reticulate`

## simPop

The `simPop` workflow starts by declaring the household identifier, household size, strata, and population status. It then simulates household structure, categorical variables, and continuous income variables.

```{r}
#| label: generate-simpop

inp <- specifyInput(
  data = eusilcP,
  hhid = "hid",
  hhsize = "hsize",
  strata = "region",
  population = TRUE
)

simPop_data <- simStructure(
  data = inp,
  method = "direct",
  basicHHvars = c("age", "gender")
)

simPop_data <- simCategorical(
  simPop_data,
  additional = c("citizenship"),
  method = "ranger",
  nr_cpus = 1
)

simPop_data <- simCategorical(
  simPop_data,
  additional = c("ecoStat"),
  method = "ranger",
  nr_cpus = 1
)

simPop_data <- simContinuous(
  simPop_data,
  additional = c("IncomeP"),
  method = "multinom",
  nr_cpus = 1
)

simPop_data <- simContinuous(
  simPop_data,
  additional = c("IncomeH"),
  method = "multinom",
  nr_cpus = 1
)

simPop_data@pop@data
simPop_data <- copy(simPop_data@pop@data)
```

## synthpop

For `synthpop`, we define a compact set of variables and use them as the visit sequence. The resulting synthetic data is converted to a `data.table` for later evaluation.

```{r}
#| label: generate-synthpop

synth_vars <- c(
  "age",
  "gender",
  "region",
  "citizenship",
  "ecoStat",
  "IncomeP",
  "IncomeH"
)

synthpop_data <- synthpop::syn(
  subset(eusilcP, select = synth_vars),
  visit.sequence = synth_vars
)

synthpop_data$syn
synthpop_data <- synthpop_data$syn
setDT(synthpop_data)
```

## Tabular ARGAN

The tabular ARGAN workflow splits the population into a household file and a person file, writes both to CSV, runs the Python generation script, and merges the generated household and person files back into one synthetic data set.

```{r}
#| label: generate-tabular-argan

eusilcP_house <- eusilcP[!duplicated(hid), .(hid, hsize, region)]
eusilcP_pers <- eusilcP[, .(hid, age, gender, citizenship, IncomeP, IncomeH)]

fwrite(eusilcP_house, file = "eusilcP_house.csv")
fwrite(eusilcP_pers, file = "eusilcP_pers.csv")

reticulate::source_python("TabularARGN.py")

tabularARGN_house <- fread("eusilcP_house_TabularARGAN.csv")
tabularARGN_house[, V1 := NULL]

tabularARGN_pers <- fread("eusilcP_pers_TabularARGAN.csv")
tabularARGN_pers[, V1 := NULL]

tabularARGAN_data <- merge(tabularARGN_house, tabularARGN_pers, by = "hid")
```

# Risk and Utility Evaluation

The following sections compute two forms of evaluation output:

- the share of replicated unique records for selected key variables
- RAPID disclosure-risk results for selected prediction models and error metrics

## Replicated Uniques

Replicated uniques are records that are unique in both the original and synthetic data for the selected key variables. A higher share indicates more one-to-one uniqueness reproduction and therefore a higher disclosure-risk signal.

```{r}
#| label: replicated-uniques-function

repl_uniques <- function(data_synth, data_orig, key_vars) {
  uniques_orig <- data_orig[, .(N_orig = .N), by = c(key_vars)]
  uniques_synth <- data_synth[, .(N_synth = .N), by = c(key_vars)]

  uniques_dt <- merge(uniques_orig, uniques_synth, by = c(key_vars), all = TRUE)
  uniques_dt[is.na(N_orig), N_orig := 0]
  uniques_dt[is.na(N_synth), N_synth := 0]
  uniques_dt[, rep_uniques := N_orig == 1 & N_synth == 1]

  uniques_dt
}
```

The same key variables are applied to all three synthetic data sets.

```{r}
#| label: replicated-uniques-analysis

key_vars <- c("age", "gender", "region", "IncomeP")

rep_u_simPop <- repl_uniques(
  data_synth = simPop_data,
  data_orig = eusilcP,
  key_vars = key_vars
)
rep_u_simPop[, sum(N_synth[rep_uniques]) / sum(N_synth)]

rep_u_synthpop <- repl_uniques(
  data_synth = synthpop_data,
  data_orig = eusilcP,
  key_vars = key_vars
)
rep_u_synthpop[, sum(N_synth[rep_uniques]) / sum(N_synth)]

rep_u_tabularARGAN <- repl_uniques(
  data_synth = tabularARGAN_data,
  data_orig = eusilcP,
  key_vars = key_vars
)
rep_u_tabularARGAN[, sum(N_synth[rep_uniques]) / sum(N_synth)]

res <- data.table(
  Method = c("simPop", "synthPop", "tabularARGAN"),
  repl.U = c(
    rep_u_simPop[, sum(N_synth[rep_uniques]) / sum(N_synth)],
    rep_u_synthpop[, sum(N_synth[rep_uniques]) / sum(N_synth)],
    rep_u_tabularARGAN[, sum(N_synth[rep_uniques]) / sum(N_synth)]
  )
)

print(res)
```

## RAPID Risk Assessment

RAPID is used to estimate record-level risk for predicting `IncomeP` from synthetic data. Two model specifications are compared:

- an intercept-only model
- a model using `gender`, `age`, and `region`

The analysis is restricted to records older than 14 years when modelling income.

```{r}
#| label: rapid-inputs

eusilcP[, Intercept := 1]
simPop_data[, Intercept := 1]
synthpop_data[, Intercept := 1]
tabularARGAN_data[, Intercept := 1]

vars_model <- list(
  c("Intercept"),
  c("gender", "age", "region")
)

v <- c("IncomeP")
filter_age <- 14

num_eps <- c(0.05, 500)
error_metric <- c("symmetric", "absolute")

output_results <- list()
```

The loop below runs RAPID for every combination of model specification and error metric, then stores one comparable result table for each synthesis method.

```{r}
#| label: rapid-loop

for (m in vars_model) {
  for (i in seq_along(num_eps)) {
    result_simPop <- rapid(
      original_data = eusilcP[age > filter_age],
      synthetic_data = simPop_data[age > filter_age],
      quasi_identifiers = m,
      sensitive_attribute = v,
      model_type = "rf",
      num_epsilon = num_eps[i],
      num_error_metric = error_metric[i],
      return_all_records = TRUE
    )

    dt_simPop <- as.data.table(result_simPop$risk$rows_risk_df)
    dt_simPop[, ID := .I]
    dt_simPop <- dt_simPop[, .(ID, at_risk)]
    dt_simPop[, Model := paste(m, collapse = " + ")]
    dt_simPop[, Sens_var := v]
    dt_simPop[, Error_metric := error_metric[i]]
    dt_simPop[, Method := "simPop"]

    result_synthpop <- rapid(
      original_data = eusilcP[age > filter_age],
      synthetic_data = synthpop_data[age > filter_age],
      quasi_identifiers = m,
      sensitive_attribute = v,
      model_type = "rf",
      num_epsilon = num_eps[i],
      num_error_metric = error_metric[i],
      return_all_records = TRUE
    )

    dt_synthpop <- as.data.table(result_synthpop$risk$rows_risk_df)
    dt_synthpop[, ID := .I]
    dt_synthpop <- dt_synthpop[, .(ID, at_risk)]
    dt_synthpop[, Model := paste(m, collapse = " + ")]
    dt_synthpop[, Sens_var := v]
    dt_synthpop[, Error_metric := error_metric[i]]
    dt_synthpop[, Method := "synthpop"]

    result_tabularARGAN <- rapid(
      original_data = eusilcP[age > filter_age],
      synthetic_data = tabularARGAN_data[age > filter_age],
      quasi_identifiers = m,
      sensitive_attribute = v,
      model_type = "rf",
      num_epsilon = num_eps[i],
      num_error_metric = error_metric[i],
      return_all_records = TRUE
    )

    dt_tabularARGAN <- as.data.table(result_tabularARGAN$risk$rows_risk_df)
    dt_tabularARGAN[, ID := .I]
    dt_tabularARGAN <- dt_tabularARGAN[, .(ID, at_risk)]
    dt_tabularARGAN[, Model := paste(m, collapse = " + ")]
    dt_tabularARGAN[, Sens_var := v]
    dt_tabularARGAN[, Error_metric := error_metric[i]]
    dt_tabularARGAN[, Method := "tabularARGAN"]

    output_results <- c(
      output_results,
      list(dt_simPop, dt_synthpop, dt_tabularARGAN)
    )
  }
}

output_results <- rbindlist(output_results)
```

Finally, the record-level RAPID output is aggregated to compare the mean risk indicator across method, model, and error metric.

```{r}
#| label: rapid-summary

out_rapid <- output_results[
  ,
  .(mean(at_risk)),
  by = .(Method, Model, Error_metric)
]

print(out_rapid)
```

# Suggested Outputs

The main objects produced by this workflow are:

- `simPop_data`: synthetic data from `simPop`
- `synthpop_data`: synthetic data from `synthpop`
- `tabularARGAN_data`: synthetic data from the tabular ARGAN workflow
- `res`: replicated-unique shares by synthesis method
- `out_rapid`: aggregated RAPID risk estimates
